babette: BEAUti 2, BEAST2 and Tracer for R

### 2018-05-02 article.tex
No empty line after listings
Describe installation and only exporting the BEAST2 config file

In R, the functions of a package need to be loaded in the global namespace first:

@@

\begin{lstlisting}[language=R, floatplacement=H]
out <- run(
  fasta_filenames = "anthus_aco.fas",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr(
      mean = create_mean_param(
        value = 1.0, 
        estimate = FALSE
      )
    )    
  )
)
\end{lstlisting}

Our initial motivation to create \verb;babette; 
was that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be 
jointly estimated with the phylogeny and other parameters. 
It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
configuration file (\cite{fix_starting_tree}), which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.
This is how to specify a fixed crown age with \verb;babette;:

\begin{lstlisting}[language=R, floatplacement=H]
out <- run(
  fasta_filenames = "anthus_aco.fas",
  posterior_crown_age = 15
)
\end{lstlisting}

\verb;babette; also supports node dating. Like BEAUti, one
can specify Most Recent Common Ancestor ('MRCA') priors.
An MRCA prior allows to specify taxa having a common ancestor,
including a distribution for the date of that ancestor.
An application of such a prior is to specify a (close to) fixed
crown age, by including all taxa and assuming a tight 
normal distribution around the desired crown age. 
Within \verb;babette;, this is achieved as such:

\begin{lstlisting}[language=R, floatplacement=H]
posterior <- run(
  fasta_filenames = "anthus_aco.fas",
  mrca_priors = create_mrca_prior(
    taxa_names = get_taxa_names("anthus_aco.fas"),
    alignment_id = get_alignment_id("anthus_aco.fas"),
    mrca_distr = create_normal_distr(
      mean = create_mean_param(value = 15.0, estimate = FALSE),
      sigma = create_sigma_param(value = 0.01, estimate = FALSE)
    )
  )
)
\end{lstlisting}

\verb;babette; allows for the same core functionality as Tracer,
in which it works on the values of the parameter estimates sampled
in the BEAST2 run. This is called the "trace" (hence the name).
The start of the trace is usually discarded, as an MCMC 
algorithm (such as used by BEAST2) first has to converge to
its equilibrium. The start of the trace, called the "burn-in", 
will be removed, because its parameter estimates are not 
representative. By default, Tracer discards the first 10\% of all 
the parameter estimates. 
To remove a 20\% burn-in from all parameter estimates 
in \verb;babette;, the following code can be used:

@@

\begin{lstlisting}[language=R, floatplacement=H]
plot_densitree(phylos = out$anthus_aco_trees)
\end{lstlisting}

>>>>

\verb;babette; can be installed easily with the \verb;devtools; package (\cite{devtools}):

\begin{lstlisting}[language=R, floatplacement=H]
library(devtools)
install_github("richelbilderbeek/babette")
\end{lstlisting}
To start using \verb;babette;, load its functions in the global namespace first:

@@

\begin{lstlisting}[language=R, floatplacement=H]
out <- run(
  fasta_filenames = "anthus_aco.fas",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr(
      mean = create_mean_param(
        value = 1.0, 
        estimate = FALSE
      )
    )    
  )
)
\end{lstlisting}
Our initial motivation to create \verb;babette; 
was that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be 
jointly estimated with the phylogeny and other parameters. 
It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
configuration file (\cite{fix_starting_tree}), which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.
This is how to specify a fixed crown age with \verb;babette;:

\begin{lstlisting}[language=R, floatplacement=H]
out <- run(
  fasta_filenames = "anthus_aco.fas",
  posterior_crown_age = 15
)
\end{lstlisting}
\verb;babette; also supports node dating. Like BEAUti, one
can specify Most Recent Common Ancestor ('MRCA') priors.
An MRCA prior allows to specify taxa having a common ancestor,
including a distribution for the date of that ancestor.
An application of such a prior is to specify a (close to) fixed
crown age, by including all taxa and assuming a tight 
normal distribution around the desired crown age. 
Within \verb;babette;, this is achieved as such:

\begin{lstlisting}[language=R, floatplacement=H]
posterior <- run(
  fasta_filenames = "anthus_aco.fas",
  mrca_priors = create_mrca_prior(
    taxa_names = get_taxa_names("anthus_aco.fas"),
    alignment_id = get_alignment_id("anthus_aco.fas"),
    mrca_distr = create_normal_distr(
      mean = create_mean_param(value = 15.0, estimate = FALSE),
      sigma = create_sigma_param(value = 0.01, estimate = FALSE)
    )
  )
)
\end{lstlisting}
\verb;babette; allows for the same core functionality as Tracer,
in which it works on the values of the parameter estimates sampled
in the BEAST2 run. This is called the "trace" (hence the name).
The start of the trace is usually discarded, as an MCMC 
algorithm (such as used by BEAST2) first has to converge to
its equilibrium. The start of the trace, called the "burn-in", 
will be removed, because its parameter estimates are not 
representative. By default, Tracer discards the first 10\% of all 
the parameter estimates. 
To remove a 20\% burn-in from all parameter estimates 
in \verb;babette;, the following code can be used:

@@

\begin{lstlisting}[language=R, floatplacement=H]
plot_densitree(phylos = out$anthus_aco_trees)
\end{lstlisting}
Instead of running the full pipeline, 
\verb;babette; also allows to only create a BEAST2 configuration file.
This file can be loaded and edited by BEAUti. To create a BEAST2 configuration
file, with all settings to default, use:

\begin{lstlisting}[language=R, floatplacement=H]
create_beast2_input_file(
  input_filenames = "anthus_aco.fas",
  output_filename = "my_beast2_input.xml"
)
\end{lstlisting}
